A Simplified Plackett-Luce Likelihood Function for Rank Orders

### 2025-11-11 inst/exponential_BT.Rmd
more work

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(hyper2)
library(stringr)
set.seed(0)
```

@@

# Try 7 competitors, focal competitor comes third

```{r}

library(partitions)
perms(7) -> x
## first r=1

jj <- x[,x[3,] == 1]

@@

```{r definegetprob, cache=TRUE}
getprob <- function(n, r, x, N=1e4){
    jj <- BT(n, x)
    wanted <- paste0("p", str_pad(r ,ceiling(log10(n))))
    c(r=r, x=x, setNames(tabulate(replicate(N, which(rrace(jj) == wanted)),nbins=n), paste0("p", 1:n)))
}
rbind(
getprob(9,1,0.5,N=100),
getprob(9,2,0.5,N=100),
getprob(9,3,0.5,N=100),
getprob(9,4,0.5,N=100),
getprob(9,5,0.5,N=100),
getprob(9,6,0.5,N=100),
getprob(9,7,0.5,N=100),
getprob(9,8,0.5,N=100),
getprob(9,9,0.5,N=100)
)

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(hyper2)
library(stringr)
library("partitions")
library("magrittr")
set.seed(0)
```

@@

# Try 8 competitors, focal competitor comes third

```{r calconetoeight, cache=TRUE}
pl <- function(v){
  out <- 1
  n <- length(v)
  for(i in seq_len(n-1)){
    out <- out*v[i]
    out <- out/sum(v[seq(from=i, to=n)])
  }
  return(out)
}

f <- function(x, place, no_of_comp){
    M <- perms(no_of_comp)
    out <- rep(0, no_of_comp)
    for(i in seq_len(ncol(M))){
        out[M[place,i]] %<>% `+`(pl(x^M[,i]))
    }
    return(out)
}

n <- 8
place <- 3
x <- seq(from = 0.1, to=1, by=0.01)
L <- matrix(0, length(x), n)
for(i in seq_along(x)){
    L[i,] <- f(x[i], place, n)
}
```

```{r onetoeight}
worker <- function(...){
    matplot(log(x),log(L), type='l', lty=1,ylim=c(-4,0))
}
pdf(file = "onetoeight.pdf")
worker()
dev.off()
worker()
```

@@

```{r definegetprob, caasdf=TRUE}
getprob <- function(n, r, x, N=1e4){
    jj <- BT(n, x)
    wanted <- paste0("p", str_pad(r ,ceiling(log10(n))))
    c(r=r, x=x, setNames(tabulate(replicate(N, which(rrace(jj) == wanted)),nbins=n), paste0("p", 1:n)))
}
rbind(
getprob(9,1,0.5,N=100),
getprob(9,2,0.5,N=100),
getprob(9,3,0.5,N=100),
getprob(9,4,0.5,N=100),
getprob(9,5,0.5,N=100),
getprob(9,6,0.5,N=100),
getprob(9,7,0.5,N=100),
getprob(9,8,0.5,N=100),
getprob(9,9,0.5,N=100)
)

### 2025-11-11 inst/hankin-exponential_BT.tex
more work

The $n=4$ case is harder as we potentially have six consistent
orderings for each observation.  As an example, if the focal
competitor comes third we have

The $n=4$ case is harder as we potentially have six consistent
orderings for each observation.  As an example, if the focal
competitor comes third we have

SDAdsfDSFFDSAfdsafdsff asdf dfs dfs dfs dsff dsfds fds fds sfd s fs sd
sdf sdf sdf sdf sdf sdf sdf sdfs dfsdf sdsdfsa asdf saf

@@

and so on.

@@

\subsection{Simplified Plackett-Luce likelihoods}

As noted, likelihood function~\ref{plackettluce} becomes unmanageable
with $n$ even moderately large and some form of simplification is
needed.  To that end, consider a Plackett-Luce likelihood function for
a race between a competitor of Bradley-Terry strength $a$ and
$n-1\geqslant 1$ cloned competitors, each of strength $b=1-a$.  Under
these circumstances, an observation is indexed by~$r$, the rank of the
focal competitor, and because we have $n$ competitors $1\leqslant
r\leqslant n$.  The initial field strength is $a+(n-1)b=1+(n-2)b$.  The
Plackett--Luce likelihood function~\citep{luce1959,plackett1975} for
this situation would be

>>>>

The general case with $n$ competitors is harder as we potentially have
$(n-1)!$ orderings for each observation.  Taking $n=4$ as an example,
if the focal competitor comes third we have

@@

and so on; see figure~\ref{1234third}.

@@

\subsection{The case $x=1$}

As noted, likelihood function~\ref{plackettluce} becomes unmanageable
with~$n$ even moderately large and we may gain additional insight by
considering a limiting case.  To that end, if $x=1$, all competitors
have the same Plackett-Luce strength.

To that end, consider a Plackett-Luce likelihood function for
a race between a competitor of Bradley-Terry strength $a$ and
$n-1\geqslant 1$ cloned competitors, each of strength $b=1-a$.  Under
these circumstances, an observation is indexed by~$r$, the rank of the
focal competitor, and because we have $n$ competitors $1\leqslant
r\leqslant n$.  The initial field strength is $a+(n-1)b=1+(n-2)b$.
The Plackett--Luce likelihood function~\citep{luce1959,plackett1975}
for this situation would be
